\section{Motivation}
Quantum computing is getting more and more popular these days.
\mycomment{twh}{Add some background of quantum computing here @wxz.}
blah\\
blah\\
blah\\
blah\\
blah\\
blah\\

There are three main models of quantum computing: \textit{Quantum Turing Machine}, \textit{Quantum $\lambda$-Calculus} and \textit{Quantum Circuit}, among which the third one is the most practical.
Most quantum programming languages are \textit{quantum-circuit description languages}, which means they are used to describe the architecture of quantum circuits.
The current quantum programming languages can be categorized into two categories according to their styles: functional quantum programming languages and imperative quantum programming languages.
\begin{multicols}{2}
  \begin{center}
      Functional:
  \end{center}
\begin{itemize}
    \item Qwire
    \item QML
    \item Quipper
    \item QuaFL
    \item Silq
\end{itemize}

  \columnbreak

  \begin{center}
          Imperative:
  \end{center}
\begin{itemize}
    \item QASM
    \item QCL
    \item ScaffCC
    \item Qiskit
    \item Quil
\end{itemize}
\end{multicols}

Advanced quantum programming languages can use more powerful abstract constructs and type systems to make it easier for programmers to write correct quantum programs.
In particular, we design a simple quantum programming language named $\lambda_Q$, which means $\lambda$-calculus with quantum circuit.
Its syntax consists of a traditional part, which is a simple $\lambda$-calculus, and a quantum part, whose syntax is based on Qwire, a functional quantum programming language with linear type system.
What's more, we implement a compiler from $\lambda_Q$ to QASM (Quantum Assembly Language), an imperative quantum programming language with low-level instruction sets.
The output QASM program can be run on the IBM cloud quantum machine.

The main feature of $\lambda_Q$ is that the syntax for traditional computation and quantum computation are separated.
They communicate with each other via some specific operations : quantum circuit can be \textit{abstracted} or \textit{lifted (measured)} into traditional term, and traditional term for quantum circuit can be applied to quantum bits.
Thus, the syntax of quantum circuit can use linear type system to guarantee that the Quantum Non-cloning Theorem is not violated, meanwhile the $\lambda$-calculus of the traditional part makes it easier to write quantum programs.